An investigation into the contentious phylogenetic results of the *Mixta* species

### 2021-01-26 Manuscript/Manuscript_Paper.Rmd
Update Manuscript_Paper.Rmd

---
title: "An investigation into the contentious phylogenetic results of the *Mixta* species"
author: "Kimberly Hinz"
date: "2020/12/08"
output:
  word_document:
    reference_docx: Styles1.docx
    fig_caption: true
    # toc: true
    # toc_depth: 5
bibliography: References.bib
csl: IJSEM.csl
---
```{r Packages, echo = FALSE, message = FALSE, warning=FALSE}
library("knitr")
library("tidyverse")
library("ggplot2")
library("RColorBrewer")
library("extrafont")
theme_set(theme_classic())
```
\newpage
# Abstract
*Mixta* is a recently described genus within the bacterial family *Erwiniaceae* and is closely related to *Pantoea*, *Erwinia*, and *Tatumella*. Depending on the genes and method used for phylogenetic analysis, the genus with which *Mixta* shares the most recent ancestor differs. This study aimed to determine the cause behind these contentious results and identified genes that may have resulted from horizontal gene transfer. Ten complete or partial genomes – two *Mixta* species, six species from other *Erwiniaceae* genera, and two outgroups – were retrieved from NCBI and annotated. Homologous genes were extracted yielding a dataset of 799 genes. The genes were aligned with the ClustalW algorithm, and MEGA-CC was used for model testing and computing phylogenetic trees for each gene for both nucleotide and amino acid sequences. Distance matrices were then extracted from these trees. Nucleotide and amino acid identity analyses were also done using the programming language `R`. *Pantoea* was the closest relative to the *Mixta* species in most analyses; however, results were not consistent. Some genes were also found to be more similar to other, non-*Pantoea* species. Diligence must be given to selecting genes for phylogenetic analysis and the method chosen to prevent any xenologous signal from distorting the actual relationships. Furthermore, future research should consider that different phylogenetic analyses may provide different results.

@@

This paper's objective was to determine the cause behind the *Mixta* species' contentious phylogenetic results. Ten complete or partial genomes – two for each of *Mixta*, *Pantoea*, *Erwinia*, and *Tatumella* and two outgroups – were retrieved from a public database. Homologous genes were extracted from these genomes in both the nucleotide and amino acid sequences. Distance matrices were exported from the phylogenetic trees, and these were used to determine which species were most closely related to *Mixta* for each gene. Nucleotide and amino acid identity analyses were also conducted for each gene to compare the results between evolutionary history and sequence similarity. Phylogenetic trees were then done with either one gene or multiple genes to confirm results.

# Materials and Methods
## Genomes and Gene Sets
Ten publicly available full or partial genomes of the species' type strains were retrieved from the National Center for Biotechnology Information (NCBI). Each of the four *Erwiniaceae* genera had two representatives and *Enterobacter cloacae* and *Pseudomonas syringae* are the outgroups (Table 1). *E. cloacae* belongs to the same order as *Erwiniacae* (*Enterobacterales*) and *P. syringae* belongs to the same class (*Gammaproteobacteria*). The genomes were annotated with Prokka v1.14.1 [@Seemann2014] using default parameters, and homologous genes were extracted using GET_HOMOLOGUES software package [@Contreras-Moreira2013] with the bidirectional best-hit search algorithm using default parameters. GET_HOMOLOGUES extracted 954 homologous genes between the ten species.

@@

## Distance Matrices
After alignments, model testing was done for each gene through `MEGA-CC` [@Kumar2012] for both nucleotide and amino acid sequences. `MEGA-CC` is called in a command-line terminal along with a text file listing the file pathways to all 799 genes and a MAO file detailing the analysis preferences. Default parameters were used for model testing. The output was a comma-separated values (CSV) file for each gene. These CSV files were read into `R`, and the model for phylogenetic analysis was extracted based on the lowest Bayesian Information Criterion (BIC).

A list of the unique models and the number of genes that required each model are given in Table 2 for nucleotide sequences and in Table 3 for amino acid sequences.

@@

kable(PM_NT, caption = "Table 2. The number of genes that required each phylogenetic tree model according to model testing and the lowest BIC for nucleotide sequences.")
```

@@

kable(PM_AA, caption = "Table 3. The number of genes that required each phylogenetic tree model according to model testing and the lowest BIC for amino acid sequences.")
```

For each unique model, a text file was written listing the gene file pathways of each gene that required that model. In a command line terminal, `MEGA-CC` [@Kumar2012] was called with a text file listing gene file pathways for a certain model and the corresponding MAO file. This was done for each unique model. These MAO files specified the use of a maximum likelihood statistical method, 500 bootstrap replications, the model, and rate patterns to create a phylogenetic tree for each gene. If a model required a Gamma distribution, the number of discrete Gamma categories was set to 5, which is default.

Distance matrices were extracted from each of the phylogenetic trees and saved as Excel files. These files were read into `R` and the genetic distances between both of the *Mixta* species and all of the sequences were extracted for further analysis in `R`. Tables were created in `R Markdown` using the `kable()` function (package: knitr) [@Xie2020] and figures were created using the `ggplot()` function (package: ggplot2) [@Wickham2016].

## Nucleotide and Amino Acid Identity
Individual species sequences of genes were aligned to both of the corresponding *Mixta* sequences in turn. Alignments were done in `R` using the same methods as for gene alignment of the whole gene file. The aligned sequences were then split into the individual sites using the `str_split()` function (package: stringr) [@]. Using base `R` code, a function was created to calculate the number of identical sites between the two sequences over the total number of sites to give a percentage of similar identity. Values that equal or are close to 100% are more similar to the *Mixta* sequences.

# Results and Discussion
After the two filters, 799 of of 954 homologous genes remained. The genes included those that encoded rRNA sequences, enzymes, transporters, and housekeeping genes and were scattered across the *M. calida* and *M. gaviniae* genomes.

## Genetic Distance
### Closest Relative from Genetic Distances
Genetic distance gives a numeric estimate of evolutionary change between species. The species that has the shortest genetic distance to a species of interest is considered to be closest relative. For the following analysis, model testing was done first to estimate the best model for each gene. Next, individual phylogenetic gene trees were created and it was from these that distance matrices were exported. *P. septica* was the closest relative to 66.3% and 66.5% of the *M. calida* and *M. gaviniae* genes, respectively, for the nucleotide sequences (Figure 1). In comparison, only 42.7% and 42.8% of *M. calida* and *M. gaviniae* genes, respectively, are closely related to *P. septica* for the amino acid sequences (Figure 1), but this is still the majority. These results suggest that *P. septica* is the closest relative to *Mixta* followed by *P. agglomerans*. For the amino acid sequences, instead of the *Mixta* genes being more closely related to *P. septica*, they are more closely related to the other species in this study (Figure 1). For example, the percentage of genes related to *P. agglomerans* increased from 13.9% for both of the *Mixta* species to 22.3 and 22.4% to *M. calida* and *M. gaviniae*, respectively. This may be due to a loss of phylogenetic signal when using amino acid sequences as a result of codon degeneracy and having a third of available sites.

```{r Figure1, echo = FALSE, message = FALSE, warning=FALSE, fig.cap="Figure 1. The percentage of genes (out of 799) that were most closely related to any of the eight non-*Mixta* species using genetic distances. The species with the shortest genetic distance from both of the *Mixta* species is considered to be the closest relative. Blue bars are the nucleotide sequences of the genes with light blue bars representing the percentage of *M. calida* genes and dark blue bars representing *M. gaviniae*. Green bars are the corresponding amino acid sequences with light green representing *M. calida* genes and dark blue bars representing *M. gaviniae*. Approximately 66% of *M. calida* and *M. gaviniae* genes were most closely related to *P. septica* when using nucleotide sequences in comparison to only 43% when using amino acid sequences.", fig.width=6.5}
number_relatives <- read.csv("C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Figure1_FirstRelativeMixtaNTAA.csv", 
                             stringsAsFactors = FALSE) %>%
  mutate(Species = factor(Species, levels = c("P. septica", "P. agglomerans", "E. tasmaniensis", "E. amylovora", "T. ptyseos", "T. saanichensis", 
                                              "E. cloacae", "P. syringae"), ordered = TRUE),
         Type = factor(Type, levels = c("calida_NT", "gaviniae_NT", "calida_AA", "gaviniae_AA"), ordered = TRUE))

@@

Unexpectedly, the number of *Mixta* genes that are more closely related increased from 2.9% to 6.0% when using the amino acid sequences instead of the nucleotide sequences (Figure 1). This, along with the other differences in relatedness for *Mixta* genes between nucleotide and amino acid sequences, suggests that the sequence type used in the analysis will give different results, even when using the same genes.

If genetic distance infers the same closest relative for a *Mixta* gene for both nucleotide and amino acid analyses, then these results may be more reliable. Thus, for genes where the closest relative is an *Erwinia*, *Tatumella*, or *Enterobacter* species both nucleotide and amino acid analyses, they may have resulted from a horizontal gene transfer (HGT) event rather than vertical evolution from a *Pantoea* species. No *Mixta* genes were found to be closely related to *P. syringae* for either the nucleotide or amino acid sequences (Figure 1). This confirms that *P. syringae* is a suitable outgroup for this analysis and suggests that any HGT events may be restricted to the order *Enterobacterales*.

### Best Model vs. Best Available Model
In MEGAX [@Kumar2018], the user is restricted to what models can be used when estimating genetic distance; however, the user is not restricted when estimating phylogenetic trees. Model testing in MEGAX tests all of the models that are available for phylogenetic tree analysis. Therefore, the recommended model based on lowest BIC value may not be available for genetic distance analysis. GTR (General Time Reversible) and HKY (Hasegawa-Kishino-Yano) models, as well as the invariant sites (+I) parameter, are not available for genetic distance analysis. In order to determine if the limited options would affect the percentage of each species being the closest relative to the *Mixta* genes, genetic distances were estimated for each gene for the nucleotide sequences in MEGA-CC. The model testing results were filtered to remove the models that are not available for genetic distances in MEGAX and then distance matrices were estimated for each gene using the best available model based on lowest BIC value. As a result, only four of twelve models were available with more than 87% of the genes now using either the Tamura 3-parameter with a Gamma parameter (T92_G) or the Tamura-Nei with a Gamma parameter (TN93_G) (Table 4). This is in comparison to the original 58% of genes that required these models when not restricted by availability. Additionally, the best model was not available for approximately 43% of the genes (242 genes). The mean BIC difference between the best model and the best available model was 2.65 with a range of 0 to 122.6 where 0 means that the best available model is the same as the best model. When ignoring cases where the best available model was also the best recommended model, the mean BIC difference was 13.8 with a range of 0.1 to 122.6. These difference may result in a less accurate estimation of genetic distance between species.

@@

kable(models, caption = "Table 4. The number of genes requiring each model when using phylogentic tree analysis vs genetic distance analysis in MEGAX for nucleotide sequences.")
```

The closest relatives to both *Mixta* species were extracted from the distance matrices. *P. septica* was estimated to be the closest relative to 12.6 and 14.4% more genes for *M. calida* and *M. gaviniae*, respectively, when the best available models were used (Figure 2). The next largest difference was for genes that were calculated to be most closely related to *P. agglomerans* (Figure 2). Approximately 5.5 and 7.1% less genes were most closely related to *P. agglomerans* when the best available model was used (Figure 2). Additionally, there was more disparity in the closest relatives between *> calida* and *M. gaviniae* when only the best available model was used (Figure 2). When the best model was used, the two *Mixta* typically had the same closest relative for each gene; however, this was the not the case when model options were limited to those available. Due to differences in evolutionary change estimation between models and the addition or exclusion of additional parameters, genetic distances may be inaccurately calculated when the best model is not used and such results may even suggest that sibling taxa are not as closely related as originally hypothesized.

```{r Figure2, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Figure 2. The percentage of genes (out of 799) that were most closely related to any of the eight non-*Mixta* species when using the recommended (best) model from model testing vs the restricted model (best available) options available for genetic distance analysis in MEGAX for nucleotide sequences. The species with the shortest genetic distance from both of the *Mixta* species is considered to be the closest relative. Evolutionary patterns may vary by gene; therefore, it is appropriate to conduct model testing and use the model that explains the most variation. Nevertheless, this model may not be available in some software programs or analysis methods within a software program. Pink bars represent the percentage of *Mixta* genes that are most closely related to any of the non-*Mixta* species when using the recommended model. Orange bars represent the percentage of *Mixta* genes that are most closely related to any of the non-*Mixta* species when using the best available model for genetic distance analysis in MEGAX in which GTR and HKY models and invariant sites parameter are not offered. The lighter shades represent *M. calida* and the darker shades represent *M. gavinaie*.", fig.width=6.5}
number_relatives <- read.csv("C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Figure2_FirstRelativeBestvsAvailableModelNT.csv", 
                             stringsAsFactors = FALSE) %>%
  mutate(Species = factor(Species, levels = c("P. septica", "P. agglomerans", "E. tasmaniensis", "E. amylovora", "T. ptyseos", "T. saanichensis", 
                                              "E. cloacae", "P. syringae"), ordered = TRUE),
         Type = factor(Type, levels = c("PD_calida", "PD_gaviniae", "GD_calida", "GD_gaviniae"), ordered = TRUE))

@@

When the closest relative results were compared between when using the best model vs when using the best available, about 20.0 and 22.3% of *M. calida* and *M. gaviniae* genes were estimated to have a different closest relative. Thus, for 82 *M. calida* genes and 64 *M. gaviniae* genes, the same closest relative was called despite using a different model. This may because the best model and the best available for those genes were similar and explained approximately the same amount of variance.

Each model estimates genetic distance differently. The best model explains most of the variance between sequences and gives a more accurate estimation of genetic distance. Using a model that does not explain most of the variance between sequences will estimate genetic distance less accurately which may result in related taxa (such as those from the same species) having a longer genetic distance between them than in actuality. This may also result in the opposite for taxa that are not closely related (i.e. *Pantoea* spp. and *E. cloacae*). Even though only two of the nucleotide models and one parameter were unavailable for genetic distance analysis in MEGAX, the results for closest relative to *Mixta* genes was substantially impacted. This was particularly the case between the two *Mixta* as one would expect them to have the same closest relative for most of the homologous genes. Therefore, using the best model is recommended. However, this may not be possible when estimating genetic distance using more than one gene that may require different models. Thus, it must be considered that one or more the genes in the concatenation may have a different evolutionary pattern than the others and results should be investigated further to ensure accuracy.

Even though genetic distance estimation in MEGAX did not offer all of the models and parameters as suggested by model testing, all of the models and parameters were available for phylogenetic tree analysis and genetic distance can be exported from these trees. However, this has to be done manually by opening the tree file in MEGAX and exporting the distance matrix as either an Excel or CSV file and is therefore not feasible when hundreds or thousands of genetic distance matrices are needed.

### Spatial Patterns of Closest Relatives
To assess whether there were any positional effects of the genes and spatial patterns of the closest relatives, circular plots were generated using the *Mixta* gene IDs for nucleotide sequences (Figure 3) and amino acid sequences (Figure 4). The gene IDs give the approximate location of each gene relative to the *Mixta* genome. The length and colour of a bar represents the species that was the closest relative to the *Mixta* gene at that position. Since the closest relatives were the same both *Mixta* species except for 1 and 2 genes for nucleotide and amino acid sequences, respectively, only the results for *M. calida* are shown. These genes will be discussed later on.

For the nucleotide sequences, there were no discernible patterns in closest relatives to *M. calida* genes with the exception of the seven genes that are most closely related to a *Tatumella* species (pink) which are only in the first half of the *M. calida* genome (Figure 3). Genes that are most closely related to *Pantoea* species (blue) were distributed throughout the genome (Figure 3), which further suggests that *Mixta* and *Pantoea* have a recent ancestor. Unexpectedly, genes that are most closely related to other genera, particularly to *E. cloacae* (orange) are also distributed throughout the genome (Figure 3).

```{r Figure3, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Figure 3. Closest relatives to *M. calida* genes for nucleotide sequences around the *M. calida* genome. Length and colour of the bars represent the different species to which the gene is most closely related according to genetic distance. The *M. calida* genome has 4084 genes. No distinct patterns emerged; however, genes that are most closely related to *P. septca* are the most common and occur throughout the entire genome. Therefore, *P. septica* and *Mixta* are likely to have a common recent ancestor.", fig.width=6.5, fig.height=5}
FR_nucl_calida <- read.csv("C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Figure3_Circular_NT.csv", stringsAsFactors = FALSE) %>%
  mutate(Closest_Relative = factor(Closest_Relative, 
                                   levels = c("P_septica", "P_agglomerans", "E_tasmaniensis", "E_amylovora", "T_ptyseos", 
                                              "T_saanichensis", "E_cloacae", "P_syringae"), ordered = TRUE))

ggplot(FR_nucl_calida, aes(x = ID, y = Relative_Number, fill = Closest_Relative)) +
  geom_bar(stat = "identity", position = position_dodge(), width = 15) +
  coord_polar() +
  scale_fill_brewer(palette = "Paired", 
                    labels = c("P. septica", "P. agglomerans", "E. tasmaniensis", "E. amylovora", "T. ptyseos", "T. saanichensis", "E. cloacae",
                               "P. syringae")) +
  scale_y_continuous(limits = c(0, 9), breaks = c(0, 2, 4, 6, 8)) +
  labs(fill = "Closest Relative") +
  theme_bw() +
  theme(legend.text = element_text(face = "italic"),
        axis.title = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        text = element_text(size = 12,  family = "Times New Roman"))
```

Less genes are closely related to *P. septica* when analyzing the amino acid sequences of the *M. calida* genes and more are closely related to *P. agglomerans*, the *Erwinia* species, the *Tatumella* species, and *E. cloacae* (Figure 4). Nevertheless, no obvious patterns arose in closest relatives to the *Mixta* genes (Figure 4). The *Mixta* genes that have changed their allegiance between the nucleotide and amino acid analyses may be a result of redundancy of the genetic code.

>>>>

---
title: "An investigation into the contentious phylogenetic results of the *Mixta* species"
author: "Kimberly Hinz"
date: "2020/12/08"
output:
  word_document:
    reference_docx: Styles1.docx
    fig_caption: true
    # toc: true
    # toc_depth: 5
bibliography: References.bib
csl: IJSEM.csl
---
```{r Packages, echo = FALSE, message = FALSE, warning=FALSE}
library("knitr")
library("tidyverse")
library("ggplot2")
library("RColorBrewer")
library("extrafont")
theme_set(theme_classic())
```
\newpage
# Abstract
*Mixta* is a recently described genus within the bacterial family *Erwiniaceae* and is closely related to *Pantoea*, *Erwinia*, and *Tatumella*. Depending on the genes and method used for phylogenetic analysis, the genus with which *Mixta* shares the most recent ancestor differs. This study aimed to determine the cause behind these contentious results and identified genes that may have resulted from horizontal gene transfer. Ten complete or partial genomes – two *Mixta* species, six species from other *Erwiniaceae* genera, and two outgroups – were retrieved from NCBI and annotated. Homologous genes were extracted yielding a dataset of 799 genes. The genes were aligned with the ClustalW algorithm, and `MEGA-CC` was used for model testing and computing phylogenetic trees for each gene for both nucleotide and amino acid sequences. Distance matrices were then extracted from these trees. Nucleotide and amino acid identity analyses were also done using the programming language `R`. *Pantoea* was the closest relative to the *Mixta* species in most analyses; however, results were not consistent. Some genes were also found to be more similar to other, non-*Pantoea* species. Diligence must be given to selecting genes for phylogenetic analysis and the method chosen to prevent any xenologous signal from distorting the actual relationships. Furthermore, future research should consider that different phylogenetic analyses may provide different results.

@@

This paper aims to determine the cause behind the *Mixta* species' contentious phylogenetic results. Ten complete or partial genomes – two for each of *Mixta*, *Pantoea*, *Erwinia*, and *Tatumella* and two outgroups – were retrieved from a public database. Homologous genes were extracted from these genomes in both the nucleotide and amino acid sequences. Distance matrices were exported from the phylogenetic trees, and these were used to determine which species were most closely related to *Mixta* for each gene. Nucleotide and amino acid identity analyses were also conducted for each gene to compare the results between evolutionary history and sequence similarity. Phylogenetic trees were then done with either one gene or multiple genes to confirm results.

# Materials and Methods
## Genomes and Gene Sets
Ten publicly available full or partial genomes of the species' type strains were retrieved from the National Center for Biotechnology Information (NCBI). Each of the four *Erwiniaceae* genera had two representatives, and *Enterobacter cloacae* and *Pseudomonas syringae* are the outgroups (Table 1). *E. cloacae* belongs to the same order as *Erwiniacae* (*Enterobacterales*), and *P. syringae* belongs to the same class (*Gammaproteobacteria*).

The genomes were annotated with Prokka v1.14.1 [@Seemann2014] using default parameters, and homologous genes were extracted using GET_HOMOLOGUES software package [@Contreras-Moreira2013] with the bidirectional best-hit search algorithm using default parameters. GET_HOMOLOGUES extracted 954 homologous genes between the ten species.

@@

## Distance Matrices
After alignments, model testing was done for each gene through `MEGA-CC` [@Kumar2012] for both nucleotide and amino acid sequences. `MEGA-CC` is called in a command-line terminal along with a text file listing the file pathways to all 799 genes and a MAO file detailing the analysis preferences. Default parameters were used for model testing. The output was a comma-separated values (CSV) file for each gene. These CSV files were read into `R`, and the model for phylogenetic analysis was extracted based on the lowest Bayesian Information Criterion (BIC). Twelve and twenty-three unique models were used for nucleotide (Table 2) and amino acid sequences (Table 3) were needed to create phylogenetic gene trees.

@@

kable(PM_NT, caption = "Table 2. According to model testing and the lowest BIC for nucleotide sequences, the number of genes required each phylogenetic tree model.")
```

@@

kable(PM_AA, caption = "Table 3. According to model testing and the lowest BIC for amino acid sequences, the number of genes required each phylogenetic tree model.")
```

The gene file paths were grouped according to unique models. For each model, a text file was written containing the pathways. `MEGA-CC` [@Kumar2012] was called in a command-line terminal along with a file pathways text file and the corresponding MAO file detailing the unique model. This step was repeated for each model. The MAO files specified the use of a maximum likelihood statistical method, 500 bootstrap replications, the model, and rate patterns to create a phylogenetic tree for each gene. If a model required a Gamma distribution, the number of discrete Gamma categories was set to 5, which is the default.

Distance matrices were extracted from each of the phylogenetic trees and saved as Excel files. These files were then read into `R`. The genetic distances between *Mixta* and the other species were extracted for further analysis in `R`. Tables were created in `R Markdown` using the `kable()` function (package: knitr) [@Xie2020] and figures were created using the `ggplot()` function (package: ggplot2) [@Wickham2016].

## Nucleotide and Amino Acid Identity
Individual species sequences of genes were aligned to both of the corresponding *Mixta* sequences in turn. Alignments were done in `R` using the same methods as for gene alignment of the whole gene file. The aligned sequences were then split into the individual sites using the `str_split()` function (package: stringr) [@Wickham2019]. A function was created using base `R` code to calculate the number of identical sites between the two sequences over the total number of sites to give a similar identity percentage. Values that equal or are close to 100% are more identical to the *Mixta* sequences.

# Results and Discussion
After the two filters, 799 of 954 homologous genes remained. The genes included those encoding rRNA sequences, enzymes, transporters, and housekeeping genes and were scattered across the *M. calida* and *M. gaviniae* genomes.

## Genetic Distance
### Closest Relative from Genetic Distances
Genetic distance gives a numeric estimate of evolutionary change between species. The species that has the shortest genetic distance to a species of interest is considered the closest relative. For the following analysis, model testing was done first to estimate the best model for each gene. Next, individual phylogenetic gene trees were created, and it was from these that distance matrices were extracted. *P. septica* was the closest relative to 66.3% and 66.5% of the *M. calida* and *M. gaviniae* genes, respectively, for the nucleotide sequences (Figure 1). In comparison, only 42.7% and 42.8% of *M. calida* and *M. gaviniae* genes, respectively, are closely related to *P. septica* for the amino acid sequences (Figure 1), but this is still the majority. These results suggest that *P. septica* is the closest relative to *Mixta*, followed by *P. agglomerans*. In contrast, the *Mixta* genes are more closely related to the other species in the study than *P. septica* in the amino acid analysis (Figure 1). For instance, the percentage of genes related to *P. agglomerans* increased by 8.4 and 8.5% for *M. calida* and *M. gaviniae*, respectively (Figure 1). This may be due to a loss of phylogenetic signal when using amino acid sequences caused by codon degeneracy and having a third of available sites.

```{r Figure1, echo = FALSE, message = FALSE, warning=FALSE, fig.cap="Figure 1. The percentage of genes (out of 799) that were most closely related to any of the eight non-*Mixta* species using genetic distances. The species with the shortest genetic distance from both of the *Mixta* species is considered the closest relative. Blue bars are the nucleotide sequences of the genes with light blue bars representing the percentage of *M. calida* genes and dark blue bars representing *M. gaviniae*. Green bars are the corresponding amino acid sequences with light green representing *M. calida* genes and dark blue bars representing *M. gaviniae*. Approximately 66% of *M. calida* and *M. gaviniae* genes were most closely related to *P. septica* when using nucleotide sequences compared to only 43% when using amino acid sequences.", fig.width=6.5}
number_relatives <- read.csv("C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Figure1_FirstRelativeMixtaNTAA.csv", 
                             stringsAsFactors = FALSE) %>%
  mutate(Species = factor(Species, levels = c("P. septica", "P. agglomerans", "E. tasmaniensis", "E. amylovora", "T. ptyseos", "T. saanichensis", 
                                              "E. cloacae", "P. syringae"), ordered = TRUE),
         Type = factor(Type, levels = c("calida_NT", "gaviniae_NT", "calida_AA", "gaviniae_AA"), ordered = TRUE))

@@

Unexpectedly, the number of *Mixta* genes that were more closely related to *E. cloacae* increased by 3.1% when amino acid sequences were used (Figure 1). The differences in relatedness for *Mixta* genes between nucleotide and amino acid sequences suggests that sequence type impacts the results, even when using the same genes.

If genetic distance infers the same closest relative for a *Mixta* gene for both nucleotide and amino acid analyses, these results may be more reliable. Thus, for genes where the closest relative is an *Erwinia*, *Tatumella*, or *Enterobacter* species both nucleotide and amino acid analyses, they may have resulted from a horizontal gene transfer (HGT) event rather than vertical evolution from a *Pantoea* species. No *Mixta* genes were closely related to *P. syringae* for either the nucleotide or amino acid sequences (Figure 1). This result confirms that *P. syringae* is a suitable outgroup for this analysis and suggests that any HGT events may be restricted to  *Enterobacterales*.

### Best Model vs Best Available Model
In `MEGAX` [@Kumar2018], there are fewer model options for genetic distance than for phylogenetic trees. However, model testing using this software tests all of the available models for phylogenetic tree analysis. Therefore, the recommended model based on the lowest BIC value may not be available for genetic distance analysis. GTR (General Time Reversible) and HKY (Hasegawa-Kishino-Yano) models, as well as the invariant sites (+I) parameter, are not available for genetic distance analysis. Genetic distances for the nucleotide sequences were estimated in `MEGA-CC` using the limited model options to determine if this would affect closest relative percentages. The model testing results were filtered to remove the models not available for genetic distances in `MEGAX`. Then distance matrices were estimated for each gene using the best available model based on the lowest BIC value. As a result, only four of twelve models were available with more than 87% of the genes now using either the Tamura 3-parameter with a Gamma parameter (T92_G) or the Tamura-Nei with a Gamma parameter (TN93_G) (Table 4). When not restricted by availability, only 58% of the genes required these models (Table 4). Furthermore, the best model was not available for approximately 30.3% of the genes (242 genes). The mean BIC difference between the best and best available models  was 2.65 (range: 0, 122.6). In this instance, 0 means that the best available model is the same as the best model. When these zeroes are ignored, the mean BIC difference was 13.8 (range: 0.1, 122.6). These differences may result in a less accurate estimation of genetic distance between species.

@@

colnames(models) <- c("Model", "Number of genes", "Number of genes, if available")

kable(models, caption = "Table 4. The number of genes requiring each model when using phylogenetic tree analysis vs genetic distance analysis in MEGAX for nucleotide sequences.")
```

The closest relatives to both *Mixta* species were extracted from the distance matrices. *P. septica* was estimated to be the closest relative to 12.6 and 14.4% more genes for *M. calida* and *M. gaviniae*, respectively, when the best available models were used (Figure 2). The next most considerable difference was for genes calculated to be most closely related to *P. agglomerans* (Figure 2). Approximately 5.5 and 7.1% fewer genes were most closely related to *P. agglomerans* when the best available model was used (Figure 2). Additionally, there was more disparity in the closest relatives between *M. calida* and *M. gaviniae* when only the best available model was used (Figure 2). When the best model was used, the two *Mixta* typically had the same closest relative for each gene; however, this was not the case when model options were limited to those available. Due to differences in evolutionary change estimation between models and the addition or exclusion of additional parameters, genetic distances may be inaccurately calculated when the best model is not used. Such results may even suggest that sibling taxa are not as closely related as initially hypothesized.

```{r Figure2, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Figure 2. The percentage of genes (out of 799) that were most closely related to any of the eight non-*Mixta* species when using the recommended (best) model from model testing vs the restricted model (best available) options available for genetic distance analysis in MEGAX for nucleotide sequences. The species with the shortest genetic distance from both of the *Mixta* species is considered the closest relative. Evolutionary patterns may vary by gene; therefore, it is appropriate to conduct model testing and use the model that explains the most variation. Nevertheless, this model may not be available in software programs or analysis methods within a software program. Pink bars represent the percentage of *Mixta* genes most closely related to non-*Mixta* species when using the recommended model. Orange bars represent the percentage of *Mixta* genes most closely related to any non-*Mixta* species when using the best available genetic distance model analysis in MEGAX. GTR and HKY models and invariant sites parameter were not available for this analysis. The lighter shades represent *M. calida*, and the darker shades represent *M. gaviniae*.", fig.width=6.5}
number_relatives <- read.csv("C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Figure2_FirstRelativeBestvsAvailableModelNT.csv", 
                             stringsAsFactors = FALSE) %>%
  mutate(Species = factor(Species, levels = c("P. septica", "P. agglomerans", "E. tasmaniensis", "E. amylovora", "T. ptyseos", "T. saanichensis", 
                                              "E. cloacae", "P. syringae"), ordered = TRUE),
         Type = factor(Type, levels = c("PD_calida", "PD_gaviniae", "GD_calida", "GD_gaviniae"), ordered = TRUE))

@@

About 20.0 and 22.3% of *M. calida* and *M. gaviniae* genes, respectively, had a different closest relative between the best and the best available models. Thus, of the 30.3% of genes with different models between the two analyses, 10.3% of the *M. calida* genes and 8.0% of the *M. gaviniae* genes had the same closest relative. These genes' results may be the same because the two models were similar and explained approximately the same amount of variance. Genetic distances between the species may also be sufficiently large that it does not change which species is most closely related to *Mixta*.

Each model estimates genetic distance differently. The best model explains most of the variance between sequences and gives a more accurate estimation of genetic distance. Using a model that does not explain most of the variance will estimate genetic distance less accurately, resulting in related taxa (such as those from the same species) having a longer genetic distance between them than in actuality. This may also result in the opposite for taxa that are not closely related (i.e. *Pantoea* spp. and *E. cloacae*). Even though only two nucleotide models and one parameter were unavailable for genetic distance analysis in `MEGAX`, the results for the closest relative to *Mixta* genes were substantially impacted. This was particularly the case between the two *Mixta* as one would expect them to have the same closest relative for most of the homologous genes. Therefore, using the best model is recommended. However, this may not be possible when estimating genetic distance using more than one gene that may require different models. Thus, it must be considered that one or more the genes in the concatenation may have a different evolutionary pattern than the others and results should be investigated further to ensure accuracy.

Although genetic distance estimation in `MEGAX` did not offer all of the models and parameters as suggested by model testing, all of the models and parameters were available for phylogenetic tree analysis and genetic distance can be exported from these trees. However, this has to be done manually by opening the tree file in `MEGAX` and exporting the distance matrix as either an Excel or CSV file. Therefore, it is not feasible when hundreds or thousands of genetic distance matrices are needed.

### Spatial Patterns of Closest Relatives
Circular plots were generated using the *Mixta* gene IDs for nucleotide (Figure 3) and amino acid sequences (Figure 4). These were created to assess whether there were any positional effects of the genes and spatial patterns in the closest relative. The gene IDs give the approximate location of each gene relative to the *Mixta* genome. The length and colour of a bar represent the species that was the closest relative to the *Mixta* gene at that position. Since the closest relatives were the same both *Mixta* species except for 1 and 2 genes for nucleotide and amino acid sequences, respectively, only the results for *M. calida* are shown. These genes will be discussed later on.

```{r Figure3, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Figure 3. Closest relatives to *M. calida* genes for nucleotide sequences around the *M. calida* genome. Length and colour of the bars represent the different species to which the gene is most closely related according to genetic distance. The *M. calida* genome has 4084 genes. No distinct patterns emerged; however, genes most closely related to *P. septica* are the most common and occur throughout the genome. Therefore, *P. septica* and *Mixta* are likely to have a recent common ancestor.", fig.width=6.5, fig.height=5}
FR_nucl_calida <- read.csv("C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Figure3_Circular_NT.csv", stringsAsFactors = FALSE) %>%
  mutate(Closest_Relative = factor(Closest_Relative, 
                                   levels = c("P_septica", "P_agglomerans", "E_tasmaniensis", "E_amylovora", "T_ptyseos", 
                                              "T_saanichensis", "E_cloacae", "P_syringae"), ordered = TRUE))

ggplot(FR_nucl_calida, aes(x = ID, y = Relative_Number, fill = Closest_Relative)) +
  geom_bar(stat = "identity", position = position_dodge(), width = 15) +
  coord_polar() +
  scale_fill_brewer(palette = "Paired", 
                    labels = c("P. septica", "P. agglomerans", "E. tasmaniensis", "E. amylovora", "T. ptyseos", "T. saanichensis", "E. cloacae",
                               "P. syringae"),
                    na.value = "#ff7f00") +
  scale_y_continuous(limits = c(0, 9), breaks = c(0, 2, 4, 6, 8)) +
  labs(fill = "Closest Relative") +
  theme_bw() +
  theme(legend.text = element_text(face = "italic"),
        axis.title = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        text = element_text(size = 12,  family = "Times New Roman"))
```

For the nucleotide sequences, there were no discernible patterns in closest relatives to *M. calida* genes except for the seven genes most closely related to a *Tatumella* species (pinks) which are only in the first half of the *M. calida* genome (Figure 3). Genes that are most closely related to *Pantoea* species (blue) were distributed throughout the genome (Figure 3), which further suggests that *Mixta* and *Pantoea* have a recent ancestor. Unexpectedly, genes closely related to other genera, particularly to *E. cloacae* (orange) are also distributed throughout the genome (Figure 3).

Fewer genes were closely related to *P. septica* when analyzing the amino acid sequences (Figure 4). Instead, more genes were closely related to *P. agglomerans*, the *Erwinia* species, the *Tatumella* species, and *E. cloacae* (Figure 4). Nevertheless, no apparent patterns arose in the closest relatives to the *Mixta* genes (Figure 4). The *Mixta* genes that have changed their allegiance between the nucleotide and amino acid analyses may result from the redundancy of the genetic code.
